---
title: "Practical Machine Learning Project"
author: "Aspiring Data Scientist"
date: "August 11, 2014"
output: html_document
---

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

First, needed libraries will be loaded and some functions defined.
```{r results='asis',echo=TRUE}
library(Hmisc)
library(caret)

# functions to be used later
# get number of NAs in each column of data frame
getColnas <- function(tdf) {
  apply(tdf, 2, function(x) sum(is.na(x)))
}
# return a data frame containing only the numeric columns in
#     that data frame
numerfy <- function(tdf) {
  # get numeric cols
  numcol = sapply(tdf,is.numeric)
  ans = tdf[,numcol]
  
}
# write 20, 1-character files
# submission function from instructor:
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
```


```{r results='asis',echo=TRUE}
#  Set up training data (trn) and test data (tst)
setwd("C:/Users/Jack/Documents/EducationalCourses/PML/prj")
# trn= 19622 obs 160 vars
# trclasse is "A", "B", "C", "D", or "E"
dataTrain = read.csv("pml-training.csv",stringsAsFactors=FALSE,
                     header=TRUE)
# tst = 20 obs 160 vars
dataTest = read.csv("pml-testing.csv",stringsAsFactors=FALSE,
                    header=TRUE)

# split dataTrain into training (trn) and (tst)
inTrain <- createDataPartition(y=dataTrain$classe,
                               p=0.7, list=FALSE)

trn <- dataTrain[inTrain,]
tst <- dataTrain[-inTrain,]
# make classe, the response variable, a factor
trn$classe <- factor(trn$classe)
tst$classe <- factor(tst$classe)
cat(nrow(trn),' obs training and ',nrow(tst),' rows test',
    sep='')
  

```
Now get rid of zero and near-zero variance predictors, that is
predictors will little information content and likely unusable
in prediction model.  Also, get rid of columns with many NAs.
```{r results='asis',echo=TRUE}
# get rid of zero and near-zero variance predictors
nzvCols <- nearZeroVar(trn)
nBefore=ncol(trn)
trn2 <- trn[,-nzvCols]
nAfter <- ncol(trn2)
cat('Columns reduced from ',nBefore,' to ',nAfter,'\n',sep='')
tst <- tst[,-nzvCols]

## remove cols with >500 NAs
nBefore=ncol(trn2)
badCol <- which(getColnas(trn2)>500)
trn3 <- trn2[,-badCol]
nAfter=ncol(trn3)
cat('Columns reduced from ',nBefore,' to ',nAfter,'\n',sep='')

classe = trn3$classe
trn4num=numerfy(trn3)
preProcObj = preProcess(trn4num,thresh=0.70, method='pca')
trn5 = predict(preProcObj,trn4num)

```

Now, model the data using the caret train() function.
First specify a train control function will create the model
using 10-fold (the default) cross validation, repeated 3 times.
```{r results='asis',echo=TRUE}
# set up trControl function for train()
cvCtrl <- trainControl(method = "repeatedcv", repeats = 3)
mdl1 = train(classe ~ ., data = trn5, method = "rpart",
      tuneLength = 30,
      trControl = cvCtrl)
trnPred = predict(mdl1,trn5)

```
Next, assess the results:
```{r}
# below gives accuracy of 0.6325
confusionMatrix(trnPred,classe)

```
Next, run the model agains the test data.  The accuracy of the
model against the training data is optimistic since the model was
fitted to that data.
The accuracy of the model against the test data should be worse.
The accuracy of the model against the test data is the best
available estimate of the out of sample error rate.
```{r}
# now, try out on test data
# pull off classe
classe = tst$classe
# make col names of tst same as training data before
#   it was run through pca
xnam=names(trn3)
tstfinal = tst[,xnam]
tstfinalNum = numerfy(tstfinal)
# preprocess so we get principal components
tstfinalpre = predict(preProcObj,tstfinalNum)
ans = predict(mdl1,tstfinalpre)
# below gives accuracy of 0.6066 which is an estimate
#    of out of sample error
confusionMatrix(ans,classe)

```
Finally, run the model against the test set and submit to Coursera.
```{r}
# prepare answers for submission to Coursera
# strip off classe from xnam
xnam=xnam[-59]
dataTest2 = dataTest[,xnam]
dataTestNum = numerfy(dataTest2)
dataTestpre = predict(preProcObj,dataTestNum)
ans2 = predict(mdl1,dataTestpre)
print(ans2)
pml_write_files(ans2)


```





